---
  output: github_document
---
  
  <!-- README.md is generated from README.Rmd. Please edit that file -->
  
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
n1 <- 1000
n2 <- 1000
```
# Fair-twin inspection

**fairadapt** is a package intended to remove bias from machine learning algorithms. In particular, it implements the pre-processing procedure described in [Plecko & Meinshausen, 2019](https://arxiv.org/abs/1911.06685) which transforms the data into its fair version, based on the causal graph of the data. One very useful property of **fairadapt** is the so called *fair-twin inspection*. In particular, for each instance (individual) in the dataset, the method computes attribute values the individual would have obtained in a *fair-world*.

## Fair-twin inspection on COMPAS

We show an example of fair-twin inspection on the COMPAS dataset.

We first load the data:
```{r load-data}
# fair-twin inspection for COMPAS
library(fairadapt)
root <- rprojroot::find_root(rprojroot::has_file("fairadapt.Rproj"))
data <- read.csv(file.path(root, "tests", "real-data", "compas", "compas-scores-two-years.csv"))
columns.keep <- which(names(data)
  %in% c("age", "sex", "juv_fel_count",
    "juv_misd_count", "juv_other_count", "priors_count",
    "c_charge_degree", "race", "two_year_recid")
)
data <- data[, columns.keep]
levels(data$race) <- c("Non-White", "Non-White", "White", "Non-White", "Non-White", "Non-White")
data$race <- relevel(data$race, "White")
train <- 1:5000
train.data <- data[train, ]
test.data <- data[-train, ]
```

We next construct the adjacency matrix of the causal graph and apply **fairadapt**:
```{r apply-fairadapt}
# set-up the adjacency matrix
adjacency.matrix <- array(0, dim = c(9, 9))
colnames(adjacency.matrix) <- c("age", "sex", "juv_fel_count",
  "juv_misd_count", "juv_other_count", "priors_count",
  "c_charge_degree", "race", "two_year_recid")
rownames(adjacency.matrix) <- colnames(adjacency.matrix)

# adding the edges to the matrix
adjacency.matrix[c("race", "sex", "age"), c("juv_fel_count", "juv_misd_count",
  "juv_other_count", "priors_count",
  "c_charge_degree", "two_year_recid")] <- 1
adjacency.matrix[c("juv_fel_count", "juv_misd_count", "juv_other_count"),
  c("priors_count", "c_charge_degree", "two_year_recid")] <- 1
adjacency.matrix["priors_count", c("c_charge_degree", "two_year_recid")] <- 1
adjacency.matrix["c_charge_degree", "two_year_recid"] <- 1

# apply fairadapt
transformed.data <- fairadapt::fairadapt(two_year_recid ~ ., train.data = train.data,
  test.data = test.data, protect.A = "race",
  adj.mat = adjacency.matrix)

# reconstruct the ordering of the data
transformed.data <- rbind(transformed.data[[1]], cbind(two_year_recid = 0, transformed.data[[2]]))
```

Finally, we select several individuals that are non-white and 30 years old. We wish to inspect the values their *fair-twins* would have obtained in a *fair world*:
```{r fair-twin}
# take several individuals
individuals <- c(241, 646, 807, 1425, 1470) # all Non-White, Age 30

# keep several interesting columns
rel.cols <- c("juv_fel_count", "juv_other_count", "priors_count")
res <- cbind(data[individuals, rel.cols], transformed.data[individuals, rel.cols])
names(res) <- c(rel.cols, paste0("fair-twin_", rel.cols))
res <- res[, c(1, 4, 2, 5, 3, 6)]
res
```

Note how we can analyze discrimination removal on individual level. In the COMPAS example, we can see the juvenile/prior counts individuals would have obtained had they been white. This property might be very useful for *justifying fair decisions* on an individual level.